\pkg{sdmTMB}: An \proglang{R} Package for Fast, Flexible, and User-Friendly Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields

### 2025-06-13 doc/jss-accepted-version/tex/jss4956.tex
Add files sent to JSS

\documentclass[article]{jss}

% setwd("doc/paper-jss/")
% knitr::knit("sdmTMB-paper.Rnw");tinytex::latexmk("sdmTMB-paper.tex", engine = "xelatex", clean = FALSE)
% R -e 'knitr::knit("sdmTMB-paper.Rnw");tinytex::latexmk("sdmTMB-paper.tex", engine = "xelatex", clean = FALSE)'

@@

\usepackage{changes}
% \usepackage[final]{changes}
\usepackage{changes}

@@

% TODO START FOR REVIEW ONLY
\usepackage{lineno}
\definecolor{darkgrey}{HTML}{A9A9A9}
\renewcommand\linenumberfont{\normalfont\bfseries\small\color{darkgrey}}
\modulolinenumbers[2]
\usepackage[dvipsnames]{xcolor}
\definecolor{niceblue}{HTML}{236899}
\newcommand{\rev}[1]{{\color{niceblue} #1}}
\newcommand{\Rev}[1]{\label{#1}\linelabel{#1}}
\newcommand{\lr}[1]{line~\lineref{#1}}
% TODO END FOR REVIEW ONLY

@@

\Abstract{
Geostatistical \added{spatial} or spatiotemporal data are common across scientific fields.
% and the dimensions of these data continue to grow in space and time.
However, appropriate models to analyse these data, such as generalised linear mixed effects models (GLMMs) with Gaussian Markov random fields (GMRFs), are computationally intensive and challenging for many users to implement.
Here, we introduce the \R\ package \sdmTMB, which extends the flexible interface familiar to users of \pkg{lme4}, \pkg{glmmTMB}, and \pkg{mgcv} to include spatial and spatiotemporal latent GMRFs using an SPDE-(stochastic partial differential equation) based approach.
SPDE matrices are constructed with \pkg{fmesher} and estimation is conducted via maximum marginal likelihood with \TMB\ or via Bayesian inference with \pkg{tmbstan} and \pkg{rstan}.
We describe the model and explore case studies that illustrate \sdmTMB's flexibility in implementing penalised smoothers, non-stationary processes (time-varying and spatially varying coefficients), hurdle models, cross-validation and anisotropy (directionally dependent spatial correlation).
% Additional functionality includes break-point effects, out-of-sample cross validation, and priors.
Finally, we compare the functionality, speed, and interfaces of related software, demonstrating that \sdmTMB\ can be an order of magnitude faster than \INLA.
We hope \sdmTMB \deleted{'s accessible interface} will help open this useful class of models to a wider field of geostatistical analysts.
}

@@

% \widowpenalty1000
% \clubpenalty1000

\begin{document}

\linenumbers

<<preliminaries, echo=FALSE, results='hide', include=FALSE, cache=FALSE>>=
library(knitr)
opts_chunk$set(engine = 'R', tidy = FALSE)
options(prompt = "R> ", continue = "+  ", width = 70, useFancyQuotes = FALSE)
@

<<main-setup, include=FALSE, cache=FALSE>>=
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  cache = TRUE,
  autodep = TRUE,
  fig.width = 7,
  fig.asp = 0.618,
  fig.pos = "ht",
  fig.align = "center",
  cache.comments = TRUE,
  dev = "pdf",
  dpi = 140,
  optipng = "-strip all",
  highlight = FALSE
)
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
options(prompt = "R> ", continue = "+  ", width = 72, useFancyQuotes = FALSE)
opts_chunk$set(prompt = TRUE)
options(replace.assign = TRUE, width = 72, prompt = "R> ")
knitr::render_sweave()
@

<<libraries, echo=FALSE, warning=FALSE, message=FALSE>>=
library(ggplot2)
library(dplyr)
library(sdmTMB)
@

<<check-version, echo=FALSE, cache=FALSE, eval=TRUE>>=
if (packageVersion("sdmTMB") < "0.6.0") {
  stop("Please install a version of sdmTMB >= 0.6.0.", call. = FALSE)
}
@

@@

Geostatistical generalized linear mixed effects models (GLMMs) with spatially correlated random effects constitute an appropriate class of models for such data \citep{rue2005gmrf, diggle2007, cressie2011, thorsonkristensen2024}.
Similarly to how random intercepts can account for correlation among groups, spatial or spatiotemporal random effects can account for unmeasured variables that cause observations to be correlated in space or both space and time.
A common approach to modelling these spatial effects is with Gaussian random fields (GRFs), where the random effects describing the spatial patterning are assumed to be drawn from a multivariate normal distribution, constrained by covariance functions such as the exponential or Matérn \citep{cressie1993, chiles1999, diggle2007}.
% TODO: decide on using GRF throughout

@@

<<matern-range, warning=FALSE, message=FALSE, fig.width=9, out.width="\\textwidth", fig.asp=0.33, fig.align='center', fig.cap="Example Gaussian random fields for two range values. The range describes the distance at which spatial correlation decays to $\\approx 0.13$ in coordinate units (i.e., the distance at which two points are effectively independent). Panel (a) shows a shorter range than panel (b), which results in a ``wigglier'' surface. Panel (c) shows the Mat\\'ern function for these two range values. The dashed horizontal line shows a correlation of 0.13.">>=
predictor_dat <- expand.grid(
  x = seq(0, 1, length.out = 100),
  y = seq(0, 1, length.out = 100),
  year = seq_len(6)
)
sim_mesh <- make_mesh(predictor_dat, xy_cols = c("x", "y"), cutoff = 0.01)
s1 <- sdmTMB_simulate(
  formula = ~1,
  data = predictor_dat,
  mesh = sim_mesh,
  range = 0.2,
  phi = 0.1,
  sigma_O = 0.2,
  seed = 1,
  B = 0
)
sim_g1 <- ggplot(s1, aes(x, y, fill = mu)) +
  geom_raster(show.legend = FALSE) +
  scale_fill_viridis_c(option = "C") +
  coord_fixed(expand = FALSE) +
  theme_light() +
  # theme(axis.title = element_blank()) +
  ggtitle("(a) Range = 0.2") +
  labs(x = "X", y = "Y")

s2 <- sdmTMB_simulate(
  formula = ~1,
  data = predictor_dat,
  mesh = sim_mesh,
  range = 0.6,
  phi = 0.1,
  sigma_O = 0.2,
  seed = 1,
  B = 0
)
sim_g2 <- ggplot(s2, aes(x, y, fill = mu)) +
  geom_raster(show.legend = FALSE) +
  scale_fill_viridis_c(option = "C") +
  coord_fixed(expand = FALSE) +
  theme_light() +
  # theme(axis.title = element_blank()) +
  ggtitle("(b) Range = 0.6") +
  labs(x = "X", y = "Y")

x <- seq(0, 1, length.out = 200)
r <- seq(0.2, 1, 0.2)
r <- c(0.2, 0.6)
df <- data.frame(
  x = rep(x, length(r)),
  range = rep(r, each = length(x))
)
matern <- function(h, sigma = 1, kappa, nu = 1) {
  ret <- (sigma^2/(2^(nu - 1) * gamma(nu))) *
    ((kappa * abs(h))^nu) *
    besselK(kappa * abs(h), nu)
  ret
  ret[x == 0] <- sigma^2
  ret
}
blues <- RColorBrewer::brewer.pal(length(r) + 1, "Blues")[-1]
df$cor <- matern(df$x, kappa = sqrt(8) / df$range)
sim_g3 <- ggplot(df, aes(x, cor, col = as.factor(range), group = as.factor(range))) +
  geom_line() +
  theme_light() +
  xlab("Distance") +
  ylab("Correlation") +
  labs(colour = "Range") +
  coord_cartesian(expand = FALSE, ylim = c(0, 1)) +
  scale_colour_manual(values = blues) +
  geom_hline(yintercept = 0.13, col = "grey50", lty = 2) +
  scale_x_continuous(breaks = r) +
  theme(legend.position = "inside", legend.position.inside = c(0.7, 0.7)) +
  ggtitle("(c) Matérn correlation function")

cowplot::plot_grid(sim_g1, sim_g2, sim_g3, align = "h", ncol = 3L)
@

@@

In practice, the above models can become considerably more complex by, for example, letting coefficients vary though time, letting coefficients vary through space \citep{hastie1993} as random fields, or adding random intercepts by grouping factors (Figure~\ref{fig:diagram}).
All these components are additive in link space.
Adopting the notation ``main'' for main effects, ``tvc'' for time-varying coefficients, and ``svc'' for spatially varying coefficients, a more complex model could be written as
\[
\begin{aligned}
\mu_{\bm{s},t} &=
g^{-1} \left( \bm{X}^{\mathrm{main}}_{\bm{s},t} \bm{\beta} +
\bm{X}^{\mathrm{tvc}}_{\bm{s},t} \bm{\gamma}_t +
\bm{X}^{\mathrm{svc}}_{\bm{s},t} \zeta_{\bm{s}} +
\alpha_g +
O_{\bm{s},t} +
\omega_{\bm{s}} +
\epsilon_{\bm{s},t} \right).
\end{aligned}
\]
where the $\bm{X}$ represent model matrices, $\bm{\gamma}_t$ represents a vector of coefficients that are constrained to vary through time as random walks or AR(1) processes, $\zeta_{\bm{s}}$ represents spatially varying coefficients following a random field, $\alpha_g$ represents IID random intercepts by group $g$ ($\alpha_g \sim \operatorname{Normal} \left(0, \sigma_\alpha^2 \right)$), $O_{\bm{s},t}$ represents an offset variable \citep[][p.~206]{mccullagh1989} (e.g., log sampling effort), and $\omega_{\bm{s}}$ and $\epsilon_{\bm{s},t}$ represent spatial and spatiotemporal intercept random fields as before (Figure~\ref{fig:diagram}).
We demonstrate these model components in Sections~\ref{pcod}, \ref{dogfish}, and \ref{svc-owls}.
% TODO: add MVN(0, sigma)s

\begin{figure}[htbp]
\centering
\includegraphics[width=\textwidth]{../../figs/diagram}
\caption[]{
Components of an \sdmTMB\ model with illustrations, descriptions, examples, notation, and example code.
An \sdmTMB\ model can be built from any combination of the process components (first six rows) plus an observation component (last row).
The examples are from an ecology context, but the model can be fit to any spatially referenced point data.
Notation: We refer to design matrices as $\bm{X}$.
The indexes $\bm{s}$, $t$, and $g$ index spatial coordinates, time, and group, respectively.
The $\sigma$ and $\bm{\Sigma}$ symbols represent standard deviations and covariance matrices, respectively.
All other symbols refer to the  described model components (e.g., $\bm{\beta}$ and $\bm{\omega}$ refer to a vector of main effects and spatial random field deviations, respectively).
Note that \code{s()} denotes a smoother as in \pkg{mgcv} \citep{wood2017a}, \code{breakpt()} denotes a breakpoint ``hockey-stick'' shape \citep[e.g.,][]{barrowman2000}, \code{(1|g)} denotes a random intercept by group \code{g}, and \code{\textasciitilde 0} is used in an \R\ formula to omit the intercept.}\label{fig:diagram}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=3.4in]{flowchart}
\caption[]{
  Description of the model fitting procedure in \sdmTMB.
}\label{fig:flowchart}
\end{figure}

@@

Once the user makes a call to \fct{sdmTMB}, input data structures for a \TMB\ model are constructed internally (Figure~\ref{fig:flowchart}).
If needed, data structures required to implement penalized smoothers are constructed with \fct{smooth2random} from \pkg{mgcv} \citep{wood2017a}, similarly to \pkg{gamm4} \citep{wood2020} and \pkg{brms} \citep{brms}.
\fct{sdmTMB} formats data, establishes parameter starting values, and constructs an objective function with derivatives based on a compiled \proglang{C++} template written for \TMB.
The objective function returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016} and efficient using sparse-matrix computation provided by \pkg{Matrix} \added{\citep{Matrix}} as \added{an} interface to the \pkg{Eigen} library in \proglang{C++}.
% TODO Jim:  I recommend adding a citation to package Matrix ( Bates D, Maechler M, Jagan M (2023). _Matrix: Sparse and Dense Matrix Classes and Methods_. R package version 1.6-4, <https://CRAN.R-project.org/package=Matrix>.)), as central dependency of TMB, given the detailed description of the dependency stack. If really motivated, you could also add the reference for Eigen (Guennebaud G, Jacob B, et al. (2010). “Eigen v3.” http://eigen.tuxfamily.org), although it's less clear what citation to use.
The negative marginal log likelihood is minimized via the non-linear optimization routine \fct{stats::nlminb} \citep{gay1990} in \R\ with additional optimization carried out via a Newton optimization step to find the Hessian using \Rev{A6} \fct{stats::optimHess} \citep{gay1990, r2024}.
Random effect values are returned at values that maximize the likelihood conditional on the fixed effects at their maximum marginal likelihood (i.e., plug-in or empirical Bayes estimates); however,
it is also possible to apply an ``epsilon'' estimator \citep{thorson2016bias}, which corrects for bias arising from the variance and skewness of random effects when calculating an estimator as a non-linear transformation or random effects (see Section \ref{index-bias-correct} for an example).
Standard errors on all parameters and derived quantities---including those involving random effects---are calculated using the generalized delta method \citep{kristensen2016, zheng2021}.
After rapid model exploration with maximum likelihood, one can optionally pass an \sdmTMB\ model to the \R\ package \pkg{rstan} \citep{carpenter2017, rstan} via \pkg{tmbstan} \citep{monnahan2018} to sample from the joint posterior distribution for Bayesian inference.

% \textbf{TODO: generalized delta method! and \citep{zheng2021}}

A fitted model summary can be viewed with \fct{print} or \fct{summary} and a set of basic ``sanity'' checks can be run with \fct{sanity}.
\fct{tidy} returns parameter estimates in standard data frame formats similar to \pkg{broom} \citep{broom}.
\Rev{A7}\added{Other} standard methods are also available such as \fct{fixef}, \fct{confint}, and \fct{vcov}.

@@

A variety of model evaluation tools are available.
A \fct{residuals} method calculates various types of residuals.
\added{The default is a form of randomized quantile \citep{dunn1996} or probability integral transform \citep{smith1985} residuals.}
New observations can be simulated from a fitted model with \fct{simulate} or new data can be simulated without a fitted model (``de novo'') with \fct{sdmTMB\_simulate}.
\fct{sdmTMB\_cv} facilitates cross validation.

% Derived quantities can be calculated on the fitted model.
% TODO: more COG refs
% Common quantities of interest include summarizing shifts in the center of gravity (\fct{get\_cog}) \citep[e.g.,][]{thorson2016cog}, or tracking total densities through time (\fct{get\_index}).

% The \fct{get\_index} function generates a weighted average of densities for the prediction grid, integrating across space; for prediction grids with equally sized cells, each cell will contribute equally to the total average, however cells may be of unequal size.

% Data may be simulated with or without a fitted model (\fct{simulate.sdmTMB}, \fct{sdmTMB\_simulate} respectively).
% Estimation is done via maximum marginal likelihood with the \fct{sdmTMB} function, and the resulting fitted model can be used to make predictions (\fct{predict.sdmTMB}) or calculate residuals (\fct{residuals.sdmTMB}), derive additional quantities of interest, and simulate Bayesian posterior distributions.
% Common quantities of interest may include summarizing shifts in the center of gravity (\fct{get\_cog}) \citep[e.g.,][]{thorson2016cog}, or tracking total densities through time (\fct{get\_index}).
% The \fct{get\_index} function generates a weighted average of densities for the prediction grid, integrating across space; for prediction grids with equally sized cells, each cell will contribute equally to the total average, however cells may be of unequal size.

@@

% TODO: describe this in more detail and how it could apply beyond ecology;
% maybe a derived variables subsection? Expanded this sentence a bit -- COGs are
% pretty common in other fields, but I don't know about indices -- removed
% `ecology` from the description

% \section{Model validation and selection}\label{model-validation-and-selection}

% Validation and selection of state-space models is challenging, particularly when
% using the Laplace approximation \citep{thygesen2017a}. We provide several
% approaches to assist this process: (1) The (marginal) Akaike Information Criterion
% \citep[AIC,][]{akaike1974} can be calculated from the marginal likelihood with \fct{AIC}, although AIC
% has well-documented biases with mixed-effects models \citep{liang2008}. (2)
% Alternatively, k-fold cross validation with \fct{sdmTMB\_cv} can be used
% with user-specified or randomly chosen folds for model selection or to evaluate
% goodness of fit according to user-calculated criteria (e.g., mean squared error,
% area under the curve). (3) An \sdmTMB\ model can be passed to the
% \pkg{tmbstan} package \citep{monnahan2018} to sample from the joint posterior
% with Stan \citep{carpenter2017}, evaluate the accuracy of the Laplace
% approximation, or perform posterior predictive checks (see
% \code{?extract\_mcmc}). (4) The \fct{residuals} method by default returns
% randomized quantile \citep{dunn1996} or probability integral transform (PIT)
% \citep{smith1985} residuals. For state-space models, these residuals have known
% statistical issues with the Laplace approximation \citep{thygesen2017a} but are
% quick to calculate. A version that uses Markov chain Monte Carlo (MCMC) sampling
% of the random effects to avoid this issue \citep{rufener2021} is recommended but
% slower (\code{?residuals.sdmTMB}). Simulation-based residuals \citep{dharma}
% are also possible. (5) The \fct{simulate.sdmTMB} method can simulate from
% fitted models and the \fct{sdmTMB\_simulate} function facilitates
% simulating data without starting from a fitted model. Models can be fit to these
% simulated data to ensure identifiability, evaluate bias and precision in
% parameter estimation, or evaluate the consequences of model misspecification.

% Jim:  If this section is re-added, you could also list the \fct{TMB::checkConsistency}
% which evaluates the accuracy (and provides a bias-correction estimator) for the
% Laplace approximation.

@@

<<sdmTMB-install, eval=FALSE, echo=TRUE>>=
install.packages("sdmTMB")
@

@@

<<sdmTMB-install-suggest, eval=FALSE, echo=TRUE>>=
install.packages("sdmTMB", dependencies = TRUE)
@

@@

<<remotes, eval=FALSE, echo=TRUE>>=
install.packages("pak")
pak::pkg_install("pbs-assess/sdmTMB", dependencies = TRUE)
@

@@

\added{
Users willing to replace the default \R\ \pkg{BLAS} (Basic Linear Algebra Subprograms) \citep{blas}
% and \pkg{LAPACK} \citep{lapack}
library with an optimized version \citep[e.g., \pkg{openBLAS};][]{openblas}
can expect up to an order of magnitude increase in model fitting speed for complex models.
Suggestions are included in the package README file.
}

@@

<<libs, warning=FALSE, message=FALSE, echo=TRUE, cache=FALSE>>=
library(sdmTMB)
library(dplyr)
library(ggplot2)
@

<<setoptions, echo=FALSE>>=
options(
  pillar.print_max = 3,
  pillar.print_min = 3,
  pillar.advice = FALSE,
  pillar.width = 80
)
options(width = 80)
@

<<libs-extras, cache=FALSE, echo=FALSE>>=
theme_set(theme_light())
options(ggplot2.continuous.fill = "viridis")
@

<<pcod-head, echo=TRUE>>=
select(pcod, lat, lon, X, Y, depth, present)
@

@@

<<pcod-utms-eval, echo=TRUE, eval=FALSE>>=
pcod <- add_utm_columns(pcod, c("lon", "lat"), units = "km")
@

@@

<<dog-binomial-mesh, results='hide', message=FALSE, warning=FALSE, echo=TRUE>>=
mesh_pcod <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 8)
@

@@

We can visualize the mesh object with the associated plotting method (Figure~\ref{fig:dog-binomial-mesh2}).
Our mesh has 563 (\code{mesh\_pcod2\$mesh\$n}) vertices.
\added{Mesh complexity has a large influence} on the speed of fitting these models.

% https://haakonbakkagit.github.io/btopic104.html
% https://becarioprecario.bitbucket.io/spde-gitbook/index.html

<<dog-binomial-mesh2, results='hide', message=FALSE, warning=FALSE, echo=TRUE, fig.cap="SPDE mesh (lines) combined with the trawl survey observations (points). The locations where lines intersect are referred to as ``vertices'' or ``knots''. Finer meshes will be slower to fit but generally increase the accuracy of the SPDE approximation, to a point. A greater degree of control over the mesh construction can be achieved by using \\pkg{fmesher} or \\proglang{R}-\\pkg{INLA} directly and supplying the object to \\code{make\\_mesh()}.", fig.width=4.5, fig.asp=1, out.width="3in">>=
mesh_pcod2 <- make_mesh(
  pcod,
  xy_cols = c("X", "Y"),
  fmesher_func = fmesher::fm_mesh_2d_inla,
  cutoff = 8,
  max.edge = c(10, 40),
  offset = c(10, 40)
)
plot(mesh_pcod2)
@

@@

<<pcod-fit, echo=TRUE>>=
fit_bin_rf <- sdmTMB(
  present ~ poly(log(depth), 2),
  data = pcod,
  mesh = mesh_pcod2,
  spatial = "on",
  family = binomial(link = "logit")
)
@

@@

<<pcod-fit-off, echo=TRUE>>=
fit_bin <- update(fit_bin_rf, spatial = "off")
@

@@

<<pcod-eg1-sanity, eval=TRUE, echo=TRUE, message=FALSE>>=
sanity(fit_bin_rf)
@

@@

<<pcod-bin-summary, echo=TRUE>>=
summary(fit_bin_rf)
@

@@

<<pcod-eg1-tidy-re, eval=TRUE, echo=TRUE>>=
tidy(fit_bin_rf, effects = "ran_pars", conf.int = TRUE)
@

@@

<<owl-plot-basic, echo=TRUE, eval=FALSE>>=
ggplot(snow, aes(X, Y)) + geom_point(aes(colour = nao_effect))
@

>>>>

\documentclass[article]{jss}\usepackage[]{graphicx}\usepackage[dvipsnames]{xcolor}
% maxwidth is the original width if it is less than linewidth
% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

\usepackage{Sweave}

@@

% \usepackage{changes}
\usepackage[final]{changes}
\usepackage{changes}

@@

@@

\Abstract{
Geostatistical spatial or spatiotemporal data are common across scientific fields.
% and the dimensions of these data continue to grow in space and time.
However, appropriate models to analyse these data, such as generalised linear mixed effects models (GLMMs) with Gaussian Markov random fields (GMRFs), are computationally intensive and challenging for many users to implement.
Here, we introduce the \R\ package \sdmTMB, which extends the flexible interface familiar to users of \pkg{lme4}, \pkg{glmmTMB}, and \pkg{mgcv} to include spatial and spatiotemporal latent GMRFs using an SPDE-(stochastic partial differential equation) based approach.
SPDE matrices are constructed with \pkg{fmesher} and estimation is conducted via maximum marginal likelihood with \TMB\ or via Bayesian inference with \pkg{tmbstan} and \pkg{rstan}.
We describe the model and explore case studies that illustrate \sdmTMB's flexibility in implementing penalised smoothers, non-stationary processes (time-varying and spatially varying coefficients), hurdle models, cross-validation and anisotropy (directionally dependent spatial correlation).
% Additional functionality includes break-point effects, out-of-sample cross validation, and priors.
Finally, we compare the functionality, speed, and interfaces of related software, demonstrating that \sdmTMB\ can be an order of magnitude faster than \INLA.
We hope \sdmTMB \deleted{'s accessible interface} will help open this useful class of models to a wider field of geostatistical analysts.
}

@@

\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}

@@

Geostatistical generalized linear mixed effects models (GLMMs) with spatially correlated random effects constitute an appropriate class of models for such data \citep{rue2005gmrf, diggle2007, cressie2011, thorsonkristensen2024}.
Similarly to how random intercepts can account for correlation among groups, spatial or spatiotemporal random effects can account for unmeasured variables that cause observations to be correlated in space or both space and time.
A common approach to modelling these spatial effects is with Gaussian random fields (GRFs), where the random effects describing the spatial patterning are assumed to be drawn from a multivariate normal distribution, constrained by covariance functions such as the exponential or Matérn \citep{cressie1993, chiles1999, diggle2007}.

@@

\begin{Schunk}
\begin{figure}[ht]

{\centering \includegraphics[width=\textwidth]{Figures/matern-range-1}

}

\caption[Example Gaussian random fields for two range values]{Example Gaussian random fields for two range values. The range describes the distance at which spatial correlation decays to $\approx 0.13$ in coordinate units (i.e., the distance at which two points are effectively independent). Panel (a) shows a shorter range than panel (b), which results in a ``wigglier'' surface. Panel (c) shows the Mat\'ern function for these two range values. The dashed horizontal line shows a correlation of 0.13.}\label{fig:matern-range}
\end{figure}
\end{Schunk}

@@

In practice, the above models can become considerably more complex by, for example, letting coefficients vary though time, letting coefficients vary through space \citep{hastie1993} as random fields, or adding random intercepts by grouping factors (Figure~\ref{fig:diagram}).
All these components are additive in link space.
Adopting the notation ``main'' for main effects, ``tvc'' for time-varying coefficients, and ``svc'' for spatially varying coefficients, a more complex model could be written as
\[
\begin{aligned}
\mu_{\bm{s},t} &=
g^{-1} \left( \bm{X}^{\mathrm{main}}_{\bm{s},t} \bm{\beta} +
\bm{X}^{\mathrm{tvc}}_{\bm{s},t} \bm{\gamma}_t +
\bm{X}^{\mathrm{svc}}_{\bm{s},t} \zeta_{\bm{s}} +
\alpha_g +
O_{\bm{s},t} +
\omega_{\bm{s}} +
\epsilon_{\bm{s},t} \right).
\end{aligned}
\]
where the $\bm{X}$ represent model matrices, $\bm{\gamma}_t$ represents a vector of coefficients that are constrained to vary through time as random walks or AR(1) processes, $\zeta_{\bm{s}}$ represents spatially varying coefficients following a random field, $\alpha_g$ represents IID random intercepts by group $g$ ($\alpha_g \sim \operatorname{Normal} \left(0, \sigma_\alpha^2 \right)$), $O_{\bm{s},t}$ represents an offset variable \citep[][p.~206]{mccullagh1989} (e.g., log sampling effort), and $\omega_{\bm{s}}$ and $\epsilon_{\bm{s},t}$ represent spatial and spatiotemporal intercept random fields as before (Figure~\ref{fig:diagram}).
We demonstrate these model components in Sections~\ref{pcod}, \ref{dogfish}, and \ref{svc-owls}.

\begin{figure}[htbp]
\centering
\includegraphics[width=\textwidth]{Figures/diagram}
\caption[]{
Components of an \sdmTMB\ model with illustrations, descriptions, examples, notation, and example code.
An \sdmTMB\ model can be built from any combination of the process components (first six rows) plus an observation component (last row).
The examples are from an ecology context, but the model can be fit to any spatially referenced point data.
Notation: We refer to design matrices as $\bm{X}$.
The indexes $\bm{s}$, $t$, and $g$ index spatial coordinates, time, and group, respectively.
The $\sigma$ and $\bm{\Sigma}$ symbols represent standard deviations and covariance matrices, respectively.
All other symbols refer to the  described model components (e.g., $\bm{\beta}$ and $\bm{\omega}$ refer to a vector of main effects and spatial random field deviations, respectively).
Note that \code{s()} denotes a smoother as in \pkg{mgcv} \citep{wood2017a}, \code{breakpt()} denotes a breakpoint ``hockey-stick'' shape \citep[e.g.,][]{barrowman2000}, \code{(1|g)} denotes a random intercept by group \code{g}, and \code{\textasciitilde 0} is used in an \R\ formula to omit the intercept.}\label{fig:diagram}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=3.4in]{Figures/flowchart}
\caption[]{
  Description of the model fitting procedure in \sdmTMB.
}\label{fig:flowchart}
\end{figure}

@@

Once the user makes a call to \fct{sdmTMB}, input data structures for a \TMB\ model are constructed internally (Figure~\ref{fig:flowchart}).
If needed, data structures required to implement penalized smoothers are constructed with \fct{smooth2random} from \pkg{mgcv} \citep{wood2017a}, similarly to \pkg{gamm4} \citep{wood2020} and \pkg{brms} \citep{brms}.
\fct{sdmTMB} formats data, establishes parameter starting values, and constructs an objective function with derivatives based on a compiled \proglang{C++} template written for \TMB.
The objective function returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016} and efficient using sparse-matrix computation provided by \pkg{Matrix} \citep{Matrix} as an interface to the \pkg{Eigen} library in \proglang{C++}.
The negative marginal log likelihood is minimized via the non-linear optimization routine \fct{stats::nlminb} \citep{gay1990} in \R\ with additional optimization carried out via a Newton optimization step to find the Hessian using A6 \fct{stats::optimHess} \citep{gay1990, r2024}.
Random effect values are returned at values that maximize the likelihood conditional on the fixed effects at their maximum marginal likelihood (i.e., plug-in or empirical Bayes estimates); however,
it is also possible to apply an ``epsilon'' estimator \citep{thorson2016bias}, which corrects for bias arising from the variance and skewness of random effects when calculating an estimator as a non-linear transformation or random effects (see Section \ref{index-bias-correct} for an example).
Standard errors on all parameters and derived quantities---including those involving random effects---are calculated using the generalized delta method \citep{kristensen2016, zheng2021}.
After rapid model exploration with maximum likelihood, one can optionally pass an \sdmTMB\ model to the \R\ package \pkg{rstan} \citep{carpenter2017, rstan} via \pkg{tmbstan} \citep{monnahan2018} to sample from the joint posterior distribution for Bayesian inference.

A fitted model summary can be viewed with \fct{print} or \fct{summary} and a set of basic ``sanity'' checks can be run with \fct{sanity}.
\fct{tidy} returns parameter estimates in standard data frame formats similar to \pkg{broom} \citep{broom}.
Other standard methods are also available such as \fct{fixef}, \fct{confint}, and \fct{vcov}.

@@

A variety of model evaluation tools are available.
A \fct{residuals} method calculates various types of residuals.
The default is a form of randomized quantile \citep{dunn1996} or probability integral transform \citep{smith1985} residuals.
New observations can be simulated from a fitted model with \fct{simulate} or new data can be simulated without a fitted model (``de novo'') with \fct{sdmTMB\_simulate}.
\fct{sdmTMB\_cv} facilitates cross validation.

@@

@@

\begin{Schunk}
\begin{Sinput}
R> install.packages("sdmTMB")
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> install.packages("sdmTMB", dependencies = TRUE)
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> install.packages("pak")
R> pak::pkg_install("pbs-assess/sdmTMB", dependencies = TRUE)
\end{Sinput}
\end{Schunk}

@@

Users willing to replace the default \R\ \pkg{BLAS} (Basic Linear Algebra Subprograms) \citep{blas}
% and \pkg{LAPACK} \citep{lapack}
library with an optimized version \citep[e.g., \pkg{openBLAS};][]{openblas}
can expect up to an order of magnitude increase in model fitting speed for complex models.
Suggestions are included in the package README file.

@@

\begin{Schunk}
\begin{Sinput}
R> library(sdmTMB)
R> library(dplyr)
R> library(ggplot2)
\end{Sinput}
\end{Schunk}

\begin{Schunk}
\begin{Sinput}
R> select(pcod, lat, lon, X, Y, depth, present)
\end{Sinput}
\begin{Soutput}
# A tibble: 2,143 x 6
    lat   lon     X     Y depth present
  <dbl> <dbl> <dbl> <dbl> <dbl>   <dbl>
1  52.3 -130.  446. 5793.   201       1
2  52.3 -130.  446. 5800.   212       1
3  52.4 -130.  449. 5802.   220       0
# i 2,140 more rows
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> pcod <- add_utm_columns(pcod, c("lon", "lat"), units = "km")
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> mesh_pcod <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 8)
\end{Sinput}
\end{Schunk}

@@

We can visualize the mesh object with the associated plotting method (Figure~\ref{fig:dog-binomial-mesh2}).
Our mesh has 563 (\code{mesh\_pcod2\$mesh\$n}) vertices.
Mesh complexity has a large influence on the speed of fitting these models.

\begin{Schunk}
\begin{Sinput}
R> mesh_pcod2 <- make_mesh(
+    pcod,
+    xy_cols = c("X", "Y"),
+    fmesher_func = fmesher::fm_mesh_2d_inla,
+    cutoff = 8,
+    max.edge = c(10, 40),
+    offset = c(10, 40)
+  )
R> plot(mesh_pcod2)
\end{Sinput}
\begin{figure}[ht]

{\centering \includegraphics[width=3in]{Figures/dog-binomial-mesh2-1}

}

\caption[SPDE mesh (lines) combined with the trawl survey observations (points)]{SPDE mesh (lines) combined with the trawl survey observations (points). The locations where lines intersect are referred to as ``vertices'' or ``knots''. Finer meshes will be slower to fit but generally increase the accuracy of the SPDE approximation, to a point. A greater degree of control over the mesh construction can be achieved by using \pkg{fmesher} or \proglang{R}-\pkg{INLA} directly and supplying the object to \code{make\_mesh()}.}\label{fig:dog-binomial-mesh2}
\end{figure}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> fit_bin_rf <- sdmTMB(
+    present ~ poly(log(depth), 2),
+    data = pcod,
+    mesh = mesh_pcod2,
+    spatial = "on",
+    family = binomial(link = "logit")
+  )
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> fit_bin <- update(fit_bin_rf, spatial = "off")
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> sanity(fit_bin_rf)
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> summary(fit_bin_rf)
\end{Sinput}
\begin{Soutput}
Spatial model fit by ML ['sdmTMB']
Formula: present ~ poly(log(depth), 2)
Mesh: mesh_pcod2 (isotropic covariance)
Data: pcod
Family: binomial(link = 'logit')

                     coef.est coef.se
(Intercept)             -0.44    0.42
poly(log(depth), 2)1   -75.72    9.30
poly(log(depth), 2)2  -109.79    9.27

Matérn range: 41.43
Spatial SD: 1.69
ML criterion at convergence: 1034.744

See ?tidy.sdmTMB to extract these values as a data frame.
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> tidy(fit_bin_rf, effects = "ran_pars", conf.int = TRUE)
\end{Sinput}
\begin{Soutput}
# A tibble: 2 x 5
  term    estimate std.error conf.low conf.high
  <chr>      <dbl>     <dbl>    <dbl>     <dbl>
1 range      41.4      9.38     26.6      64.6 
2 sigma_O     1.69     0.233     1.29      2.22
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> ggplot(snow, aes(X, Y)) + geom_point(aes(colour = nao_effect))
\end{Sinput}
\end{Schunk}
